Finally tagless, partially evaluated

### 2007-04-06 tagless/tagless-final.tex
Mostly fiddling with the title

\title{Finally Tagless Staged Interpreters for Simpler Typed Languages}
%Staged typed type-preserving tagless ``interpreters'', \emph{finally}
%Finally tagless, simply typed, partially evaluated
\subtitle{\raisebox{-2.5em}[0pt][0pt]{\normalsize\it\begin{minipage}{30em}
    It should also be possible to define languages with a highly
    refined syntactic type structure. Ideally, such a treatment should
    be metacircular, in the sense that the type structure used in the
    defined language should be adequate for the defining language.
    \hfill\rm\citep{reynolds-definitional}
\end{minipage}}}

\begin{comment}
\authorinfo{Jacques Carette}
           {McMaster University}
           {carette@mcmaster.ca}
\authorinfo{Oleg Kiselyov}
           {FNMOC}
           {oleg@pobox.com}
\authorinfo{Chung-chieh Shan}
           {Rutgers University}
           {ccshan@cs.rutgers.edu}
\end{comment}
\authorinfo{}{}{}

@@

We turn to a more general approach to CBV CPS: a CPS transformer that
turns any implementation of |Symantics| into a CPS interpreter, whether
it is an evaluator.  This functor on interpreters performs a textbook
CPS transformation on the object language.
\begin{code}
module CPST(S: Symantics) = struct
  let int i = S.lam (fun k -> S.app k (S.int i))
  let add e1 e2 = S.lam (fun k ->
    S.app e1 (S.lam (fun v1 ->
    S.app e2 (S.lam (fun v2 ->
    S.app k (S.add v1 v2))))))
  let lam f = S.lam (fun k -> S.app k (S.lam
    (fun x -> f (S.lam (fun k -> S.app k x)))))
  let app e1 e2 = S.lam (fun k -> 
    S.app e1 (S.lam (fun f ->
    S.app e2 (S.lam (fun v ->
    S.app (S.app f v) k)))))
  let fix f = S.fix (fun self -> (f self))
end
\end{code}
The (abbreviated) code explicitly maps CPS interpretations to
(direct) interpretations performed by 
the base interpreter~|S|.

@@

With 
|CPST| instantiated by the compiler~|C| above,
|T.test1| gives
\begin{code}
.<fun x_5 ->
  (fun x_2 -> x_2 (fun x_3 x_4 -> x_4 x_3))
  (fun x_6 -> (fun x_1 -> x_1 true)
              (fun x_7 -> x_6 x_7 x_5))>.
\end{code}
This output is the na\"{\i}ve CPS transformation of $(\fun{x}x)\True$,
containing several apparent $\beta$-redexes.  To eliminate these
redexes, we simply change~|T| to instantiate |CPST| with |P| instead.
\begin{code}
{P.st = Some <fun>;
 P.dy = .<fun x_5 -> x_5 true>.}
\end{code}

@@

Our initial motivation came from several papers 
\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} which used
embedded interpreters as a central example to justify looking 
into advanced type systems.  While we admire all this technical work,
we were not convinced that their 
motivating example actually needed all this machinery.

>>>>

\title{Finally tagless, partially evaluated}
\subtitle{Tagless Staged Interpreters for Simpler Typed Languages}
%Staged typed type-preserving tagless ``interpreters'', \emph{finally}

\begin{comment}
\authorinfo{Jacques Carette}
           {McMaster University}
           {carette@mcmaster.ca}
\authorinfo{Oleg Kiselyov}
           {FNMOC}
           {oleg@pobox.com}
\authorinfo{Chung-chieh Shan}
           {Rutgers University}
           {ccshan@cs.rutgers.edu}
\end{comment}
\authorinfo
{\smash{\normalsize\it\begin{minipage}{30em}
    It should also be possible to define languages with a highly
    refined syntactic type structure. Ideally, such a treatment should
    be metacircular, in the sense that the type structure used in the
    defined language should be adequate for the defining language.
    \hfill\rm\citep{reynolds-definitional}
\end{minipage}}}{}{}

@@

We turn to a more general approach to CBV CPS: a CPS transformer that
turns any implementation of |Symantics| into a CPS interpreter, whether
it is an evaluator.  This functor on interpreters performs a textbook
CPS transformation on the object language.
\begin{code}
module CPST(S: Symantics) = struct
  let int i = S.lam (fun k -> S.app k (S.int i))
  let add e1 e2 = S.lam (fun k ->
    S.app e1 (S.lam (fun v1 ->
    S.app e2 (S.lam (fun v2 ->
    S.app k (S.add v1 v2))))))
  let lam f = S.lam (fun k -> S.app k (S.lam
    (fun x -> f (S.lam (fun k -> S.app k x)))))
  let app e1 e2 = S.lam (fun k -> 
    S.app e1 (S.lam (fun f ->
    S.app e2 (S.lam (fun v ->
    S.app (S.app f v) k)))))
  let fix f = S.fix (fun self -> (f self))
end
\end{code}
This (abbreviated) code explicitly maps CPS interpretations to
(direct) interpretations performed by 
the base interpreter~|S|.

@@

With 
|CPST| instantiated by the compiler~|C| above,
|T.test1| gives
\begin{code}
.<fun x_5 ->
  (fun x_2 -> x_2 (fun x_3 x_4 -> x_4 x_3))
  (fun x_6 -> (fun x_1 -> x_1 true)
              (fun x_7 -> x_6 x_7 x_5))>.
\end{code}
This output is a na\"{\i}ve CPS transformation of $(\fun{x}x)\True$,
containing several apparent $\beta$-redexes.  To eliminate these
redexes, we just change~|T| to instantiate |CPST| with |P| instead.
\begin{code}
{P.st = Some <fun>;
 P.dy = .<fun x_5 -> x_5 true>.}
\end{code}

@@

Our initial motivation came from several papers 
\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} that used
embedded interpreters to justify
advanced type systems.  While we admire all this technical work,
we were not convinced that their 
motivating example actually needed all this machinery.
